---
title: "OCN683_Programming_HW3"
author: "ʻAʻaliʻi Kelling"
date: "2025-01-31"
output: html_document
---
## **Homework 3 - Some Programming Practice**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Instruction ** 
The goal of this assignment is to have you practice some basic R programming / datawrangling. The attached file, “CRCP_Reef_Fish_Surveys_Hawaii_expanded.csv”, includes survey data of reef fish on Hawaiʻi island collected by NOAA (yellow circles in the map
above = survey location, see here). I have slightly modified the data (added zeros for when a species it not observed at a location, removed some columns we will not be using). 

```{r, echo=FALSE}
#load R packages
library(dplyr)
library(tidyr)
library(tidyverse)
library(psych)
library(car)
library(janitor) #cleaning data function
library(here) #enables easy file referencing 
library(rstatix) #descriptive stats
library(ggeffects) #effects plot
library(ggplot2)
library(emmeans)
library(ggpubr)
library(fishualize)

fish <- read_csv("CRCP_Reef_Fish_Surveys_Hawaii_expanded_posted.csv")
str(fish)
fish <- fish %>%
  mutate_at(vars(commonname, taxonname, species, depth_bin), 
            as.factor)

```

## **(1) QUESTION** Let’s look at some basic summary statistics. 

In total there are 209 fish species encountered and counted in these surveys. Identify the top twelve species in terms of these three statistics:**(1) mean abundance**, **(2) maximum abundance**, **(3) standard deviation of abundance**. In other words:
1. Which species are most abundant on average? 
2. Which reach the highest abundances, and which have the most variable abundances?

```{r}
#Descriptive stats
summary(fish)
sapply(fish, class)
sppcounts <- summary(fish$commonname)

#?dplyr
#View(fish)

#Running descriptive stats using the dplyr package
sppstats <- fish %>%
  group_by(commonname) %>%
  summarise(
    meanabund = mean(count),
    maxabund = max(count),
    sdabund = sd(count)
  )
sppstats

#Identifying the top 12 fish species
top12_meanabund <- sppstats %>% arrange(desc(meanabund)) %>% slice (1:12)
top12_maxabund <- sppstats %>% arrange(desc(maxabund)) %>% slice (1:12)
top12_sdabund <- sppstats %>% arrange(desc(sdabund)) %>% slice (1:12)

#RESULTS
top12_meanabund 
top12_maxabund 
top12_sdabund 


top12_meanabund$commonname[1:12]

top12spp <- c("Vanderbilt's chromis", "Agile chromis", "Brown surgeonfish",
  "Spotted surgeonfish", "Yellow tang", "Common bluestripe snapper",
  "Arc-eye hawkfish", "Hawaiian bicolor chromis", "Whitebar surgeonfish",
  "Saddle wrasse", "Mackerel scad", "Striated wrasse") 

#Making a table that only looks at specified species
top12 <- fish %>%
  select(commonname, scientific_name, count, depth, depth_bin) %>%
  filter(commonname %in% top12spp) %>%  # Use species list
  mutate(commonname = factor(commonname, levels = top12spp)) %>%  # set the levels (orders)
  arrange(commonname)

top12
str(top12)

```

## Instructions
To make these calculations, you’ll want to focus on the ‘count’ column (number of fish observed at a location), as well as the ‘commonname’ or ‘taxonname’ columns. The latter two are nearly identical, you can choose which you prefer to use. To do these calculations in base R, tapply() or aggregate() will be useful, or to use the ‘tidyverse’ you can look into
the summarize() function in the package dplyr.


## Answer
Based on our results, Vanderbilt's chromis was the most abundant species on average, with a mean abundance of 30 individuals per survey in reef fish surveys around Hawaiʻi Island. This species was observed nearly twice as frequently as the next most abundant species, Agile chromis, which had a mean abundance of 19 individuals per survey.

The Common bluestripe snapper (Taʻape) recorded the highest single-survey abundance, reaching 600 individuals in one instance. This aligns with its known behavior as a social schooling species, often forming dense aggregations, which explains its ability to reach such high numbers in specific surveys.

Interestingly, despite being the most abundant species on average, Vanderbilt's chromis also exhibited the highest variability in abundance, with a standard deviation (SD) of 53. This suggests that while they can be found in large numbers in certain areas, they may be completely absent in others. This pattern makes sense, as chromis are schooling fish, meaning their distribution is often patchy—they cluster in preferred habitats but may be scarce or absent in less suitable areas. Their high variability in abundance reflects their tendency to form dense aggregations in some locations while being entirely absent in others.


## **(2) QUESTION** Now let’s visualize how the abundances of the most common species vary with depth, to get a sense for whether species have different depth niches. 

**Using the top twelve species based on mean abundance:**
**Plot a scatterplot of count vs. depth (the column is named ‘depth’).** Add a smoother to help visualize mean count vs. depth, put all twelve species on a single plot (with twelve panels), and make sure each panel is titled using the species name. Furthermore, do not arrange the panels in alphabetical order — instead, arrange them in order of mean abundance, so that the most abundant species is in the top left panel, and remaining panels are in order of mean abundance. The patterns in abundance will be easier to see if you transform the counts or the y-axis, for example with a square root, because the counts are very skewed. 


I would like you to make this twelve-paneled figure in two ways. **First, write a ‘for’ loop in which you iterate your plotting code for the twelve species.** The function scatter.smooth() is a simple one for making a scatterplot with a smoother. **Second, use ggplot() to do the same thing.** The functions geom_point(), geom_smooth(), and facet_wrap() will be helpful. The ggplot way of doing this may be slightly easier (once you’ve figured out the ggplot syntax), but I would like you to do it both ways so that I know that you know how to write a loop, which is a very general and important programming technique.

**What are your conclusions from this visual inspection of the data?** Make sure that the patterns in the plots are visible. If they are not, you may need to adjust the figure dimensions in Markdown.


#Tranforming Data
```{r}

# Create histogram for fish count
fishcount <- hist(top12$count) 
abline(v=mean(top12$count),col="blue")
abline(v=median(top12$count),col="red")
fishcount #Left Skewed

# Create histogram for fish depth
fishdepth <- hist(top12$depth) 
abline(v=mean(top12$depth),col="blue")
abline(v=median(top12$depth),col="red")
fishdepth #Not bad

#square root transformation on count and depth
top12 <- top12 %>%
  mutate(
    count_sqrt = sqrt(count),
    depth_sqrt = sqrt(depth),
    count_log = log10(count + 1))


#Looking at histogram again with transformed values for count and depth

# Create histogram for transformed fish count
fishcount_log <- hist(top12$count_log) 
abline(v=mean(top12$count_log),col="blue")
abline(v=median(top12$count_log),col="red")
fishcount_log #Left Skewed but looks better than it was before

# Create histogram for transformed fish depth
fishdepth_sqrt <- hist(top12$depth_sqrt) 
abline(v=mean(top12$depth_sqrt),col="blue")
abline(v=median(top12$depth_sqrt),col="red")
fishdepth_sqrt #Not bad

```


#Using Loops
```{r}
#View(top12)

# Set up a 3x4 multi-panel layout for 12 species
par(mfrow = c(3,4))  
# Loop through the top 12 species
for (i in 1:12) {  
  # Subset data for the current species
  species_data <- subset(top12, commonname == top12spp[i])  

    # Scatter plot with smoother
    scatter.smooth(
      x = species_data$depth_sqrt,   #transformed depth
      y = species_data$count_log,  #transformed count  
      main = top12spp[i],   # Title with species names 
      xlab = "Depth (m)",   
      ylab = "Log Transformed Fish Count",   
      col = "blue",  
      lpars = list(col = "red", lwd = 2)  # Smoother line in red  
    )  
}


# Scatterplot of abundance vs. depth for top 12 species using ggplot
ggplot(top12, aes(x = depth_sqrt, y = count_sqrt, color = commonname)) +
  geom_point(alpha = 0.5) +  # Scatterplot with transparency
  geom_smooth(method = "loess", se = FALSE) +  # Smoother for each species
  geom_smooth(method = "lm", linetype = "solid", color = "black", se = FALSE) +  # Add regression line
  facet_wrap(~ commonname, scales = "free") +  # Facet by species
  labs(
    title = "Abundance vs. Depth for Top 12 Species",
    x = "log transformed Depth (m)",
    y = "log transformed Fish Count",
    color = "Species"
  ) +
  theme_minimal()


```
**What are your conclusions from this visual inspection of the data?** Make sure that the patterns in the plots are visible. If they are not, you may need to adjust the figure dimensions in Markdown.

The raw fish count data exhibited a right-skewed distribution, which is expected for count-based data. To better visualize patterns, both count and depth were log-transformed, which helped normalize the data and improve interpretability. Although the transformed data still did not appear as visually clean as expected, it provided a clearer comparison than the raw distributions.

From the scatterplots, we observe distinct depth-abundance trends among species:
- Species showing increasing abundance with depth: Agile chromis, Common bluestripe snapper, Hawaiian bicolor chromis, and Striated wrasse exhibited positive trends, meaning their abundance increased with greater depth.
- Species showing decreasing abundance with depth: Brown surgeonfish, Yellow tang, Whitebar surgeonfish, and Saddle wrasse showed a negative trend, with abundance declining as depth increased.
- Species with no strong depth-related trends: Vanderbilt’s chromis, Spotted surgeonfish, Arc-eye hawkfish, and Mackerel scad displayed relatively constant distributions across depths, with no clear increasing or decreasing trends.

Based on visual inspection, Surgeonfish species appear to be more abundant in shallower depths, while some wrasse and chromis species are more abundant at greater depths and the lack of strong trends in some species may indicate that their distribution is influenced by other ecological factors rather than depth by itself

## **(3) QUESTION ** Finally, let’s focus on just the top 5 most abundant species.
Make a new plot that shows abundance vs. depth for the top 5 species, including smoothers, but this time put all of the species on the same scatterplot and distinguish them with different colors. This time you don’t need to use two different approaches to make the plot — one approach will suffice. What is your interpretation of this plot? Fit a linear model that tests whether these 5 species have different depth niches. Based on residual diagnostic plots of this model, and a plot of the fitted effects, do you think a linear model is a good approach for testing this question? If not, why not? 


```{r}

#First 5 species
top5spp <- top12spp[1:5]  

top5 <- top12 %>%
  filter(commonname %in% top5spp)


# Scatterplot of abundance vs. depth for top 5 species
ggplot(top5, aes(x = depth_sqrt, y = count_sqrt, color = commonname)) +
  geom_point(alpha = 0.5) +  # Scatterplot with transparency
  geom_smooth(method = "loess", se = FALSE) +  # Smoother for each species
  labs(
    title = "Abundance vs. Depth for Top 5 Species",
    x = "Log-Transformed Depth (m)",
    y = "Log-Transformed Fish Count",
    color = "Species"
  ) +
  theme_minimal()



```
## **What is your interpretation of this plot?**
The scatterplot shows the relationship between fish abundance and depth for the top five most abundant species, with each species represented by a distinct color. While the data appears somewhat noisy, general trends can still be observed.

Both chromis spp (Vanderbilt’s chromis and Agile chromis) exhibit higher abundances at deeper depths (3–5 m) compared to the other species. Their distribution follows a curved pattern, suggesting that their abundance peaks at certain depths rather than increasing or decreasing linearly. This could indicate preferred depth ranges rather than a simple positive or negative relationship with depth.

In contrast, surgeonfish spp (Brown surgeonfish, Spotted surgeonfish, and Yellow tang) are distributed across all depths but tend to have lower abundances overall compared to the chromis species. Surgeonfish abundance appears to have a negative relationship with depth, as they are more frequently observed in shallower depths (~1–3 m) and decline in abundance as depth increases.

Overall, these trends suggest that chromis species are more commonly associated with deeper reef zones, while surgeonfish are more evenly distributed but with a preference for shallower waters. However, the curved patterns seen in some species indicate that depth preferences may not be strictly linear, highlighting the potential need for non-linear models to better understand these relationships.



## Linear Model
```{r}
# Fit a linear model testing if species have different depth niches
fish_model <- lm(depth_sqrt ~ commonname + count_sqrt, data = top5)

# Model summary
summary(fish_model)

par(mfrow = c(2,2))  # Arrange plots in a 2x2 grid
plot(fish_model)  # Standard diagnostic plots

# Predict fitted values
top5$predicted_depth <- predict(fish_model)

# Plot the fitted effects
ggplot(top5, aes(x = commonname, y = predicted_depth, color = commonname)) +
  geom_point(size = 4) +
  labs(title = "Predicted Depth for Each Species",
       x = "Species",
       y = "Predicted √(Depth)") +
  theme_minimal()


```

## **Answer**
Of the top five most abundant species, the results indicate that most of them are found at significantly different depths. Agile chromis (p = 1.02e-10), Spotted surgeonfish (p = 2.53e-05), and Yellow tang (p = 0.0003) all showed statistically significant differences in the depths they occupy, suggesting that these species have distinct depth preferences. However, when looking at the relationship between fish abundance and depth, the results show no significant effect (p = 0.53). This means that the number of individuals present does not influence the depths at which these species are found, and other factors are likely shaping their distributions.

Despite these significant results, the model has a very low R-squared value (R² = 0.0117), meaning that species identity and abundance together explain only about ~1.17% of the variation in depth. This suggests that while species do exhibit depth preferences, there are likely other environmental or ecological factors driving depth variation that were not accounted for in this model. Variables such as habitat complexity, reef structure, competition, or environmental conditions could be playing a much larger role in determining where these species are found.

Given the weak R-squared value and the limited explanatory power of the model, I do not think that a linear model was the best approach for testing this question. The low explanatory power suggests that depth preferences are not linear relationships and might be better analyzed using a Generalized Additive Model (GAM) or a Generalized Linear Model (GLM), which allow for more flexibility in capturing non-linear trends. Additionally, incorporating more ecological variables into the model could provide a clearer picture of what actually influences fish depth distributions. Further analysis is needed to explore these additional factors and determine what is truly driving the observed depth variations among these species.








